
\begin{comment}

  - what is infrastructure
  - why infrastructure is soo important
    - it is the skelenton of the program
    - once infrastructure is in place, it defines what is easy to change, and what can't change without major refactor


\end{comment}


Our project, irrelevent to implementation language has a very simliar foundation to all games, requiring networking between computers playing the game, graphics engine that can draw the game, assets/sprites for in game entities.

The game's implementation will be designed with an infrastructure that is at the core. All game features will run on top of this infrastructure, having little dependency on other game features, but having complete dependency on the infastrcture. This architecture allows very module design with regards to game features, allowing new game features to be added to the implementation only a slight modification of the infrastructure to include the new feature. By designing the architecture like this, It faciliates a release based schedule.

A high level specification is now available that specifices the elemnts of the game. To bring this high level specification to reality, the game design will be broken down into 3 major releases: Alpha, Beta 1, and Beta 2. Each release specifies the game features that must be in the game. Essentially the game design is broken down into game features, with higher priority game features being specified in earlier releases.
Alpha Release is primarily focused on building the game's Infrastructure, this is the core of the game with all the back end logic that is never seen by the player. By the end of Alpha 1 the game infrastructure is mostly complete, and the game is in a state where game logic can now be added on like modules with little dependency 
Beta 1 and Beta 2 are both game feature oriented, building on top of the existing infrastructure, adding features described in the specification. These releases will heavily focus on meeting the specification's specified game features. 