\section{Conclusions}

%todo: needs proof reeding
Many successful projects experience severe problems, scheduling errors, delays and resource shortfalls, but that in no way brands the project as a failure, what truly measures the project's success is whether the the project management compensates for failure and it able to mitigate the effects. This project was carefully structured so that even under the worst possible circumstances the project could be refactored to mitigate the scope creep so the project remains true to the specification. 

The agile development cycles in conjunction with the scheduling required that each week there was a stable release featuring a new component. This model transpired to be over-ambitious, and features were not released on a weekly schedule but were instead released on a bi-weekly schedule, however it's work pointing out that this reflected an error in the development cycle length rather than a delay in development, and that releases were still stable throughout the project. Communication was as effective as communication can be expected to be; sometimes colleagues think they're agreeing when they're disagreeing, sometimes they're arguing over a point they are in fact agreeing on but incorrectly expressing, communication is and will always be an issue in all types of projects. The project's risk management provided the team insight to allow extra time for dealing with problems encountered during development, and mitigate the affect on the overall project.

Some parts of the project cannot even be assessed until long after the project is finished. Often legal or social issues don't become apparent until a long time after the project is completed, most project analysis tools don't become useful until the project is long since completed by which time lessons are lost, stakeholder perspectives change, or the project becomes an relic. What's truly important above all else is to ensure that the correct strategies have been identified, the mistakes have been documented, the teams have learned a lesson and that the experience will be applied to the next project. The team, and the majority of the stakeholders, would definitely consider the project a success.
