\section{Legal, Ethical, and Social Issues}
\label{sec:legal}

Prior to embarking on this project several potential professional issues were identified
and discussed to ensure that they were addressed appropriately.

Any software project faces issues surrounding copyright and intellectual property. During
development the team were aware that all third party libraries used in the project must
be licensed appropriately. This meant only using software with a permissive license (e.g.
Apache, BSD, or MIT licenses) and no unlicensed proprietary software. It was also important
to avoid infringing on the intellectual property of established game publishers and their
existing games. Obviously it was infeasible to review all previously published games to
check if Serenity bears great similarities to any of them. However, the project team was
careful not to directly copy any designs or unique game play characteristics of existing
games that they are familiar with.

The games industry is faced with several ethical and social issues that were thought about
carefully when designing the Serenity game. Highly controversial issues that surrounds
current games, such as racism and violence, were thought about carefully and it was ensured
that they would not be issues with the Serenity game. This is because the game is based
in a fictional world with a very unrealistic graphical representation. Another social
dilemma for game creators to consider is that of the addictiveness of their creations.
Players often invest many hours, often consecutively, in playing games and can even spend
large sums of money on pay-to-play games. The game that was developed was designed to
have short playing periods that would allow convenient times for the player to quit.
