\section{Review of Requirements}
\label{sec:reviewrequirements}

The project was successful in fulfilling most of the requirements but, due to delays during development, not all the features could be completed in time. Rather than risk not having a playable game, some of the minor features were deferred until the end of the project, and focus was shifted to the project's critical chain.

\subsection{Functional Requirements}

\begin{description}
%This section needs to be reworked to explain the success of each one compared to what was required.
  \item[FR1 Multiplayer] The specification required two player games to be supported, at present the game will theoretically support any number of players, but in practice the game dynamics would decay when the number of players exceeded the expected capacity of the selected map. The game met and exceeded the multiplayer requirement outlined in the specification.

  \item[FR2 Fleets] A core concept of the specified game was for players to be able to create and customise a fleet of ships, including an option of equipping weapons of the player's choice. Each player is now given the opportunity to customise their fleet constructing ships they wish to include in deployment and (optionally) naming ships. The requirement for including a fleet of ships was fulfilled, but some of the finer details of the requirement such as shield recharging remain incomplete.

  \item[FR3 Ship Customisation] The specification required ships to have customisable weapon slots, currently each of the fleet's ships can be constructed using one of three different hull types. Each type offers a different ship speed, health, and shield capacity. The larger the ship, the higher its health/shield capacity, and the slower its speed, so the requirement was partially met, but the players are at present unable to select weapon slots.

  \item[FR4 Resource System] The framework for the resource system was completed. However, the game mechanics for resources remains unfinished. The implementation of the remaining features would be fairly straightforward in the event that this project is continued, but from the player's perspective the requirement is unmet.

  \item[FR5 Planetary Capture] Players can capture a neutral planet if their fleet secures the area for a sufficient amount of time. Once the planet has been held securely by one player for a sufficient amount of time, the planet then falls under that player's control, which would provide them with a steady stream of resources and contribute towards a victory condition. Planets can also be captured from other players in the same way a neutral plant can be captured, except it takes the capturing player longer to secure the planet, giving an advantage to being the first player to capture a planet. This requirement was completed as required by the specification.

  \item[FR6 Fog of War] Although it was a relatively simple concept, fog of war was not completed in time since it was low priority and time was scarce. The project failed to meet this requirement entirely.

  \item[FR7 AI] An advanced artificial intelligence framework was implemented. It took orders from the player and converted them into a lower level plan to execute. Some replanning is possible, but it is not a formal piece of the framework and it could be improved. Intelligent path finding algorithms were created to get ships to take the optimal route to their planned destinations. This AI system mostly met requirements, but is a bit less powerful than originally intended. However, its flexibility should allow it to be extended in useful ways in the future.

  \item[FR8 Campaign] A campaign of battles between two players was always an `if there is enough time' feature. Unfortunately, due to time constraints a campaign mode has not been implemented yet.

  \item[FR9 Operating System Requirements] The game compiles and runs on both Mac OS X and Linux. Windows systems were not tested during development, but measures were taken to ensure the game can be supported on windows, for example there are function calls available in the sockets module to facilitate networking on Windows.

\end{description}

\subsection{Non-Functional Requirements}
\begin{description}
  \item[NFR1 Fun] The project group find the game fun, as did a small group of pre-alpha testers, but since there was no time for a formal play testing there is little data to support how \emph{fun} the game experience is. The game experience almost certainly would be improved should missing features been implemented in time; the project group even had concepts to improve gameplay beyond the scope of the specification.
    
  \item[NFR2 Short game sessions] Game session timings are not enforced by the game, and there is insufficient data available to gauge average play time, however, it seems that under normal circumstances a game should be no longer than 35 minutes if players are seriously aiming to win. The specification required the game be no longer than 35 minutes, but also didn't specify any action that needed to be taken in the time limit was reached, so the requirement can be deemed partially complete.

  \item[NFR3 Approachable] Only the host's IP address is required to play a networked game, which is typical of most standalone games. A player is also able to host and join games using an elegant GUI making it very easy to get started.

  \item[NFR4 Reliable] The game is required to be reliable enough to have a gaming session without crashing. The software was thoroughly tested to ensure it wouldn't easily crash and results show the game, other than a few minor bugs, it meets the reliability requirement.
  
  \item[NFR5 Secure] Since the game runs as with a master server controlling the entire simulation of the game the amount of cheating that is possible for a player is significantly reduced. Incoming messages are verified to be from whichever player they claim to be sent from by checking the originating IP and port. Players are also blocked from sending commands to ships that they do not actually control. However, packet forgery is still possible since there is no authentication or integrity checking at the network layer. This was briefly investigated by using a message authentication code algorithm such as SipHash,\cite{aumasson2012siphash} but it was decided that it would be better to focus efforts on completing the game logic and return to network layer packet verification at a later date.

\end{description}

%% Table in the same order as listed

%\begin{table*}[ht]
%	\begin{tabular}{p{10em} p{6em} p{12em}}
%		\toprule
%		\emph{Requirement} & \emph{Priority} & \emph{Requirement Status}\\
%		\midrule
%   Multiplayer & Vital & Fulfilled\\
%	Fleets & Vital &  Fulfilled \\
%	Ship Customisation & High & Partly Fulfilled\\
%	Resource System & Medium & Not Fulfilled \\
%	Planetary Capture & High & Fulfilled \\
%	Fog of War & Low & Not Fulfilled \\
%	AI & High & Fulfilled \\
%	Campaign &  Optional & Not Fulfilled\\
%	Operating System & High & Fulfilled\\
%	Fun & High & Fulfilled\\
%	Short game sessions & Medium & Fulfilled\\
%	Approachable & High & Fulfilled\\
%	Reliable & Vital & Fulfilled\\
%	Secure & Medium & Fulfilled\\
%	\bottomrule
%	\end{tabular}
%	\caption{Requrement fullfillment}
%	\label{tab:requrementstable}
%\end{table*}


\begin{table*}[ht]
	\begin{tabular}{p{10em} p{6em} p{12em}}
		\toprule
		\emph{Requirement} & \emph{Priority} & \emph{Requirement Status}\\
		\midrule
   Multiplayer & Vital & Fulfilled\\
	Fleets & Vital &  Fulfilled \\
	Reliable & Vital & Fulfilled\\
	
	Ship Customisation & High & Partly Fulfilled\\
   Planetary Capture & High & Fulfilled \\
   AI & High & Fulfilled \\
	
   Operating System & High & Fulfilled\\
	Fun & High & Fulfilled\\
	Approachable & High & Fulfilled\\
	Resource System & Medium & Not Fulfilled \\
	Short game sessions & Medium & Fulfilled\\
   Secure & Medium & Fulfilled\\
	Fog of War & Low & Not Fulfilled \\
	Campaign & Optional & Not Fulfilled\\

	%todo maybe someone can improve this table
	\bottomrule
	\end{tabular}
	\caption{Requirement fulfillment}
	\label{tab:requrementstable}
\end{table*}

\vspace{5mm}

As can be seen from \ref{tab:requrementstable}, only one high priority requirement remains incomplete, the only other requirements which were not met were lower priority or entirely optional anyway. The project failed to meet all the requirements laid out in the specification, but it certainly succeeded in meeting enough to meet the projects overall aim -- to produce a playable game.
