\section{Roles and Responsibilities}
\label{section:roles}

Here are described the roles and responsibilities required for the project. The holders of these responsibilities at the start of the project are shown in table \ref{tab:roles}, but it is envisaged that there are some roles that could rotate (for example Software Librarian). Some roles are to be held by a single person at a time, others may have several.

\begin{description}
     
    \item[Team Leader] Leads the project. Responsible for coordinating the project team, ensuring team members are working to the best of their ability, responsible for making decisions when the there is no clear solution to a particular problem.
      
    \item[Lead Developer] Responsible overall for technical areas of the project. Consults other developers when there is development difficulty, also responsible for directing programming style and technique.
    
    \item[Project Manager] Responsible for overseeing the human aspects of the project in general, including managing a schedule, organising meetings and collaborative development sessions. Makes decisions involving tradeoffs between project time, cost, and quality. 
     
    \item[Customer Liaison] Meets with the customer at regular intervals to discuss the project progress, outlook, and any issues which are require customer input. 
        
    \item[Analyst] Responsible for ensuring the customer requirements are addressed during planning and development stages of the project, and ensures that the solution will sufficiently address the customers needs.
     
    \item[User Manager] Responsible for communicating with playtesters and users. Finds end users to test the product in the later stages of development, and provides feedback to the project team. Prime responsibility is to identify issues which are not clearly visible from the project development perspective, but are more apparent to end users.
     
    \item[Graphic Designer] Responsible for prototyping and developing graphical design elements, such as ship sprites, terrain, maps and user interface.
     
    \item[Software Librarian] Ensures team completes documentation to a sufficient standard for long term maintenance. 
     
    \item[Line Manager] Oversees day to day development, intervenes if a developer is off track, ensuring minimal time is wasted perusing low priority work.
     
    \item[Code Reviewer] Responsible for interpreting other developer's code, checking for logical inconsistencies and familiarising themselves with the project as a whole.
     
    \item[Chairman] Responsible for coordinating meetings, ensuring all issues are resolved or at the least discussed, and that all meeting participants have a chance to voice concerns and contributions.
     
    \item[Testing and Integration Manager] Ensures the code is thoroughly tested for bugs, and discovered bugs are flagged and dealt with in reasonable time. Responsible for managing integration testing to prevent bugs occurring on the master branch.
     
    \item[Security Officer] Checks for security flaws in the product, performs security evaluations (such as penetration testing and code review) to ensure the product is sufficiently secure.
     
    \item[Composer] Composes soundtrack for the game, and produces required sound effects, voice overs, soundscapes, and other related resources.
    
    \item[Tester] Performs general code testing (e.g. unit tests, component tests).
    
    \item[Programmer] Performs the day-to-day programming specified by the line manager.
    
    \item[Gameplay Designer] Critically analysis gameplay design and experience, giving feedback to the team leader on how to improve the games's appeal.
\end{description}

\begin{table*}[h!]
	\begin{tabular}{l p{38em}}
		\toprule
		\emph{Person} & \emph{Description} \\
		\midrule
		\emph{Laith Alissa} & Project manager, Customer Liaison, User Manager, Analyst, Graphics Designer.\\[0.5em]
		\emph{Jon Cave} & Code Reviewer, Chairman, Security Officer, Testing and Integration.\\[0.5em]
		\emph{Joseph Siddall} & Software Librarian, Line Manager, Testing and Integration.\\[0.5em]
		\emph{Vic Smith} & Team Leader, Lead Developer, Composer.\\[0.5em]
		\bottomrule
		\caption[][1em]{Roles of project group members at the start of the project.}
	\end{tabular}
	\label{tab:roles}
\end{table*}