\chapter[Project Requirements]{Project Requirements}
\label{ch:reference}

\chapterepigraph{Rien n'est plus difficile, et donc plus pr\'ecieux, que d'\^etre capable de d\'ecider}{Napol\'eon Bonaparte}

\newthought{Here presented are the aims}, endpoints, controls, methods, schedule, and every relevant detail of the project; laid out unequivocally for reference before, during, and after the required work.

\section{Objectives and Deliverables}

\begin{margintable}
\vspace{12em}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}
{p{8em} p{7em}}
		\toprule
		\emph{Deliverable} & \emph{Deadline} \\
		\midrule
		Prototype Game & Dec 6th 2012\\
		Progress Poster & Dec 6th 2012\\
		Final game & April 25th 2013\\
		Report & April 25th 2013\\
		Presentation & May 7th 2013 \\
		\bottomrule
\end{tabular}
\vspace{0.5em}
\caption{Summary of Deadlines}
\end{margintable}

The first objective of the project team is to produce a playable prototype of the game specified. 
The prototype can be extremely basic, provided the playing experience gives a promising outlook for the final release. The prototype release is both an internal motivating factor and a requirement for the progress report at the end of term 1.

The progress poster is the first external deadline, and is a customer requirement for ensuring the project is on schedule. The progress poster will contain technical detail of the project progress to date, as well as screen captures of the latest running game prototype.

The final game should be completed to the requirements laid out in the specification and to amendments agreed on by the customer. An end user should be able to play a networked game with minimal configuration (no more configuration required than a typical installation and networking configuration).

The presentation and final report are the last deliverables of the project. They give the team the opportunity to deconstruct the project's progress over the academic year, and give an analysis of the finished product. The report documents the design approach taken, relevant research which influenced the project, quality standards review, testing, user manuals, and any critical decisions made during the project, and the presentation will give the project team a chance to introduce the game, and offers the customer an opportunity to question the team. 

%might need to rephrase the double and

\input{chapter/pid/requirements}
\input{chapter/pid/design_approach}
\input{chapter/pid/quality}
\input{chapter/pid/foreseeable_challenges}
\input{chapter/pid/schedule}
